Figures for Paper

### 2021-06-28 paper_figures.Rmd
alsod
Code for scrapping ALS mutations from ALSoD website + technical variates

@@

# NMD experiment on SH-SY5Y cells

@@

```{r}
####Inclusion reads by if TDP-potential####
boxplot_table = clean_data_table %>%
    mutate(across(UNC13A_3prime_leaf:UNC13A_annotated_leaf, ~ .x / library_size,.names = "{.col}_library_norm")) %>%
    filter(!tissue_clean %in% c("Choroid","Liver")) %>%
    dplyr::select(sample,participant_id,mutations,disease_group2,pathology,tissue_clean,contains("_library_norm")) %>%
    melt() %>%
    filter(grepl("_3prime|_5prime_1",variable)) %>%
    group_by(sample) %>%
    mutate(inclusion_reads = sum(value)) %>%
    ungroup() %>%
    unique() %>%
    mutate(junction_name = case_when(variable == "UNC13A_3prime_leaf_library_norm" ~ "  Novel Donor",
                                 variable == "UNC13A_5prime_1_leaf_library_norm" ~ " Short Novel Acceptor",
                                 variable == "UNC13A_5prime_2_leaf_library_norm"~ "Long Novel Acceptor")) %>%
    mutate(disease_tissue = case_when((grepl("FTLD",disease_group2) & grepl("Cortex",tissue_clean))  ~ T,
                                      (grepl("ALS",disease_group2) & grepl("Cord|Motor",tissue_clean))  ~ T,
                                      (grepl("Occipital",tissue_clean)) ~ F,
                                      (grepl("Control",disease_group2) & grepl("Cord|Cortex",tissue_clean)) ~ T,
           TRUE ~ F)) %>%
    mutate(tissue_clean = gsub("_"," ",tissue_clean))

@@

boxplot_table %>%
  filter(disease_tissue == T) %>%
  mutate(detected = inclusion_reads > 0) %>%
  dplyr::select(participant_id,disease_group2,detected) %>%
  unique() %>%
  group_by(disease_group2) %>%
  mutate(n_sample = n_distinct(participant_id)) %>%
  mutate(n_sample_detected = sum(detected)) %>%
  dplyr::select(disease_group2,n_sample,n_sample_detected) %>%
  unique() %>%
  mutate(detection_rate = n_sample_detected / n_sample) %>%
  mutate(disease_group2 = gsub("Control"," Control",disease_group2)) %>%
  mutate(detection_name = glue::glue("{disease_group2} \n ( {n_sample} )")) %>%
  ggplot() +
  geom_col(aes(x = detection_name, y = detection_rate)) +
  ggpubr::theme_pubr() +
  scale_y_continuous(lim = c(0,1),labels = scales::percent) +
  ylab("Percent of Patients \n UNC13A Cryptic Detected") +
  theme(text = element_text(size = 24)) +
  xlab("N individuals")
```

@@

>>>>

<!-- # iPSC WTC11 parent line genotype -->

<!-- ```{r confirm_no_als_ftd_wtc11 mutations} -->
<!-- library(vcf) -->
<!-- wtc11_vcf = "/Users/annaleigh/Downloads/3546dc62_AH77TTBBXX_DS-229105_GCCAAT_recalibrated.vcf.gz" -->
<!-- vcf <- read.vcfR( wtc11_vcf, verbose = FALSE ) -->
<!-- ``` -->

@@

## scatterplot UNC13A CE and UNC13B NMD

```{r}
rel_rna_cryptic_amount %>%
    filter(condition != "control") %>%
    ggplot() +
    geom_point(aes(x = cryptic_psi, y = normalized_unc13b_ir, fill = source),pch = 21,size = 4) +
    stat_cor(aes(x = unc13b_nmd_exon_psi, y = normalized_unc13b_ir)) +
    theme(legend.position = 'bottom')+
    facet_wrap(~source)

```
# NMD experiment on SH-SY5Y cells

@@

```{r Inclusion reads by if TDP-potential}
####Inclusion reads by if TDP-potential####
boxplot_table = clean_data_table %>%
    mutate(across(UNC13A_3prime_leaf:UNC13A_annotated_leaf, ~ .x / library_size,.names = "{.col}_library_norm")) %>%
    filter(!tissue_clean %in% c("Choroid","Liver")) %>%
    dplyr::select(sample,participant_id,mutations,disease_group2,pathology,tissue_clean,contains("_library_norm")) %>%
    melt() %>%
    filter(grepl("_3prime|_5prime_1",variable)) %>%
    group_by(sample) %>%
    mutate(inclusion_reads = sum(value)) %>%
    ungroup() %>%
    unique() %>%
    mutate(junction_name = case_when(variable == "UNC13A_3prime_leaf_library_norm" ~ "  Novel Donor",
                                 variable == "UNC13A_5prime_1_leaf_library_norm" ~ " Short Novel Acceptor",
                                 variable == "UNC13A_5prime_2_leaf_library_norm"~ "Long Novel Acceptor")) %>%
    mutate(disease_tissue = case_when((grepl("FTLD",disease_group2) & grepl("Cortex",tissue_clean))  ~ T,
                                      (grepl("ALS",disease_group2) & grepl("Cord|Motor",tissue_clean))  ~ T,
                                      (grepl("Occipital",tissue_clean)) ~ F,
                                      (grepl("Control",disease_group2) & grepl("Cord|Cortex",tissue_clean)) ~ T,
           TRUE ~ F)) %>%
    mutate(tissue_clean = gsub("_"," ",tissue_clean))

@@

clean_data_table %>%
  filter(disease_tissue == T) %>%
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(detected = inclusion_reads > 0) %>%
  dplyr::select(participant_id,disease_group2,detected) %>%
  unique() %>% 
  group_by(disease_group2) %>%
  mutate(n_sample = n_distinct(participant_id)) %>%
  mutate(n_sample_detected = sum(detected)) %>%
  dplyr::select(disease_group2,n_sample,n_sample_detected) %>%
  unique() %>%
  mutate(detection_rate = n_sample_detected / n_sample) %>%
  mutate(disease_group2 = gsub("Control"," Control",disease_group2)) %>%
  mutate(detection_name = glue::glue("{disease_group2} \n ( {n_sample} )")) %>%
  ggplot() +
  geom_col(aes(x = detection_name, y = detection_rate)) +
  ggpubr::theme_pubr() +
  scale_y_continuous(lim = c(0,1),labels = scales::percent) +
  ylab("Percent of Patients \n UNC13A Cryptic Detected") +
  theme(text = element_text(size = 24)) +
  xlab("N individuals")
```

## Effect of library size on CE discovery

```{r unc detection by library size  }
# library size

clean_data_table %>%
   filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  filter(disease_tissue == T) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-") ) %>% 
  ggplot(aes(x = unc13a_detected,y = log10(library_size) )) + 
  geom_point(colour = "firebrick", size = 0.5, alpha = 0.9, position = position_jitter(width = 0.2, height = 0) ) + 
  geom_boxplot(fill = NA, colour = "black", outlier.colour = NA) +
  labs(title = "Library size and UNC13A CE detection ",subtitle = "Wilcox test") + 
  facet_wrap(~ tissue + disease_group2) + 
  ggpubr::stat_compare_means(label = "p.format", label.x.npc = "center", method = "wilcox" ) +
  labs(y = "log10 library size", x = "UNC13A CE detected") +
  theme(axis.text.x = element_text(size = 15)) + 
  ggpubr::theme_pubr()
```

## Effect of UNC13A TPM onCE discovery

```{r unc detection by UNC13A tpm }
# library size

clean_data_table %>%
   filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  filter(disease_tissue == T) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-") ) %>% 
  ggplot(aes(x = unc13a_detected,y = UNC13A_TPM )) + 
  geom_point( colour = "firebrick", size = 0.5, alpha = 0.9, position = position_jitter(width = 0.2, height = 0) ) + 
  geom_boxplot(fill = NA, colour = "black", outlier.colour = NA) +
  labs(title = "UNC13A TPM and UNC13A CE detection ",subtitle = "Wilcox test") + 
  facet_wrap(~ tissue + disease_group2, scales = 'free_y') + 
  ggpubr::stat_compare_means(label = "p.format", label.x.npc = "center", method = "wilcox" ) +
  labs(y = "UNC13A expression (TPM)", x = "UNC13A CE detection") +
  theme(axis.text.x = element_text(size = 15)) +
  ggpubr::theme_pubr()

## Effect of RIN and UNC13A discovery

```{r rin discovery}
## effect of RIN?
clean_data_table %>%
  filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  filter(disease_tissue == T) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  filter(!is.na(rin)) %>%
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-") ) %>% 
  ggplot(aes(x = unc13a_detected,y = rin )) + 
  geom_point( colour = "firebrick", size = 0.5, alpha = 0.9, position = position_jitter(width = 0.2, height = 0) ) + 
   geom_boxplot(fill = NA, colour = "black", outlier.colour = NA) +
  labs(title = "Sample RIN and UNC13A CE detection ",subtitle = "Wilcox test") + 
  facet_wrap(~disease_group2 + tissue) + 
  ggpubr::stat_compare_means(label = "p.format", label.x.npc = "center" ) +
  labs(y = "RIN", x = "UNC13A CE detection") +
  theme(axis.text.x = element_text(size = 15)) +
  ggpubr::theme_pubr()

```{r unc13a tpm and ce detection}
## effect of RIN?
clean_data_table %>%
  filter(disease_tissue == T) %>% 
  filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  add_count(tissue_clean,disease_group2) %>% 
  mutate(plot_name = glue::glue("{disease_group2} \n ({n})")) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-") ) %>% 
  ggplot(aes(x = plot_name, fill = unc13a_detected,y = UNC13B_TPM )) + 
  geom_point(size = 2,show.legend = F,pch = 21, position = position_jitterdodge(dodge.width=1,jitter.width = 0.3))+
  geom_boxplot(outlier.colour = NA) +
  labs(title = "UNC13A CE detection and UNC13B TPM",subtitle = "Wilcox test") + 
  facet_wrap(~ tissue_clean,scales = 'free') + 
  ggpubr::stat_compare_means(label = "p.signif",) +
  labs(y = "UNC13B TPM", x = "UNC13A CE detection") +
  theme(axis.text.x = element_text(size = 15)) +
  ggpubr::theme_pubr()
```

## Effect of celltype composition on UNC13A CE detection

```{r}
conversion_table <- data.frame(short = c("Ast","End", "Mic", "Ex", "Oli", "Per" ),
                               long = c("Astrocytes", "Endothelial", "Microglia", "Neurons", "Oligos", "Pericytes")
  )

cell_types = fread("/Users/annaleigh/Documents/GitHub/unc13a_cryptic_splicing/data/MuSiC_residual_results.tsv")

cell_types %>% 
  left_join(clean_data_table %>%
   filter(disease_group2 %in% c("FTLD-TDP", "ALS-TDP")) %>%
  filter(disease_tissue == T) %>% 
  mutate(inclusion_reads = UNC13A_3prime_leaf + UNC13A_5prime_1_leaf) %>% 
  mutate(unc13a_detected = ifelse(inclusion_reads > 0, yes = "+", no = "-") ))

@@

# Relationship between STMN2 and UNC13B fsE PSI
## Scatter plot showing the correlation in non-log space in cortex UNC13B fsE & STMN2 CE

```{r}
clean_data_table %>%
  filter(disease_tissue == T) %>%
  filter(disease_group2 %in% c("FTLD-TDP","ALS-TDP")) %>%
  filter(junction_reads_stmn2 >= 30) %>%
  filter(junction_reads_unc13a >= 30) %>%
  mutate(junction_reads_unc13b = UNC13B_nmd_junction_5prime + UNC13B_annotated + UNC13B_nmd_junction_3prime) %>% 
  filter(junction_reads_unc13b >= 30) %>%
  # filter(UNC13A_3prime_leaf + UNC13A_5prime_1_leaf >= 2) %>%
  # filter(STMN2_cryptic_leaf >= 2) %>%
  filter(grepl("Cortex",tissue_clean)) %>%
  filter(stmn_2_cryptic_psi_leaf > 0 ) %>%
  filter(unc13a_cryptic_leaf_psi > 0) %>%
  ggplot(aes(x = unc13a_cryptic_leaf_psi, y = unc13b_nmd_psi)) +
  geom_point() +
  stat_cor(size = 8,method = "spearman",cor.coef.name = "rho") +
  geom_smooth(method = lm, se = FALSE,color = "black") +
  xlab("UNC13A CE PSI") +
  ylab("UNC13B fsE PSI") +
  ggpubr::theme_pubr() +
  theme(text = element_text(size = 18)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L))  +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1L))
```
